---
title: "Dynamic branching with raster tiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic branching with raster tiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# With the root.dir option below,
# this vignette runs the R code in a temporary directory
# so new files are written to temporary storage
# and not the user's file space.
knitr::opts_knit$set(root.dir = fs::dir_create(tempfile()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.height = 5,
  fig.width = 5
)
if (identical(Sys.getenv("NOT_CRAN", unset = "false"), "false")) {
  knitr::opts_chunk$set(eval = FALSE)
}
library(targets)
library(geotargets)
Sys.setenv(TAR_ASK="false")
```

```{r setup}
library(geotargets)
library(targets)
library(terra)
library(dplyr)
```

Computationally intensive raster operations that work in pixel-wise manner may be handled well with [dynamic branching](https://books.ropensci.org/targets/dynamic.html#about-dynamic-branching) over tiled subsets of the raster.
`tar_terra_tiles()` is a target factory that enables creating these dynamic branches so that downstream targets can iterate over them.
This is useful when, for example, loading an entire raster into memory and doing computations on it results in out of memory errors.

## Tile helpers

`geotargets` provides three helper functions that take a `SpatRaster` and output extents for tiles.

```{r}
#example SpatRaster
f <- system.file("ex/elev.tif", package="terra")
r <- rast(f)
```

`tiles_n()` is the simplest, which produces *about* `n` tiles in a grid.

```{r}
tile_n(r, 4)
```

For more control, use `tile_grid()` which allows specification of the number of rows and columns to split the raster into.

```{r}
tile_grid(r, ncol = 3, nrow = 1)
```

The third included helper is `tile_blocksize()` which tiles by file block size.
Block size is a property of raster files and is the number of pixels (in the x and y direction) that is read into memory at a time.
Tiling by multiples of block size may therefore be more efficient because only one block should need to be loaded to create each tile target.

```{r}
tile_blocksize(r, n_blocks_row = 1, n_blocks_col = 2)
```

This only works when the `SpatRaster` points to a file—in-memory rasters have no inherent block size.

```{r error=TRUE}
sources(r)
#force into memory
r2 <- r + 0
sources(r2)
#this now errors
tile_blocksize(r2)
```

## Example targets pipeline

When developing a `targets` pipeline using `tar_terra_tiles()` with `tile_blocksize()`, it's a good idea to figure out how many tiles `tile_blocksize()` will create before implementing `tar_terra_tiles()`.
We'll start by making a bigger raster to experiment with using `terra::disagg()` and making multiple layers.

```{r}
#| label: tar_script
#| echo: false
#| eval: true

targets::tar_script({
    # contents of _targets.R
    library(targets)
    library(geotargets)
    library(terra)
    geotargets_option_set(gdal_raster_driver = "COG")
    list(
        tar_target(
            raster_file,
            system.file("ex/elev.tif", package="terra"),
            format = "file"
        ),
        tar_terra_rast(
            r,
            disagg(rast(raster_file), fact = 10)
        ),
        tar_terra_rast(
            r_big,
            c(r, r + 100, r * 10, r / 2),
            memory = "transient"
        )
    )
})

```

```{r}
#| label: _targets
#| eval: false
#| echo: true
# contents of _targets.R
library(targets)
library(geotargets)
library(terra)
geotargets_option_set(gdal_raster_driver = "COG")
list(
    tar_target(
        raster_file,
        system.file("ex/elev.tif", package="terra"),
        format = "file"
    ),
    tar_terra_rast(
        r,
        disagg(rast(raster_file), fact = 10)
    ),
    tar_terra_rast(
        r_big,
        c(r, r + 100, r * 10, r / 2),
        memory = "transient"
    )
)
```

```{r}
#| label: tar_make
#| eval: true
#| echo: true

tar_make()
tar_load(r_big)
tile_blocksize(r_big)
```

Four tiles is reasonable, so we'll go with that.
Note that we have to ensure the `r_big` target is not in-memory for `tar_terra_tiles()`, so we set `memory = "transient"`.
To merge tiles back together into a single `SpatRaster` we can use

```{r}
#| label: tar_script2
#| eval: true
#| echo: false
targets::tar_script({
    # contents of _targets.R
    library(targets)
    library(geotargets)
    library(terra)
    geotargets_option_set(gdal_raster_driver = "COG")
    tar_option_set(memory = "transient")
    list(
        tar_target(
            raster_file,
            system.file("ex/elev.tif", package="terra"),
            format = "file"
        ),
        tar_terra_rast(
            r,
            disagg(rast(raster_file), fact = 10)
        ),
        tar_terra_rast(
            r_big,
            c(r, r + 100, r * 10, r / 2),
            memory = "transient"
        ),
        tar_terra_tiles(
            tiles,
            raster = r_big,
            tile_fun = tile_blocksize,
            description = "split raster into tiles"
        ),
        tar_terra_rast(
            tiles_mean,
            app(tiles, \(x) mean(x, na.rm = TRUE)),
            pattern = map(tiles),
            description = "some computationaly intensive task performed on each tile"
        ),
        tar_terra_rast(
            merged_mean,
            merge(sprc(tiles_mean)),
            description = "merge tiles into a single SpatRaster"
        )
    )
})

```

```{r}
#| label: _targets2
#| eval: false
#| echo: true
# contents of _targets.R
library(targets)
library(geotargets)
library(terra)
geotargets_option_set(gdal_raster_driver = "COG")
tar_option_set(memory = "transient")
list(
    tar_target(
        raster_file,
        system.file("ex/elev.tif", package="terra"),
        format = "file"
    ),
    tar_terra_rast(
        r,
        disagg(rast(raster_file), fact = 10)
    ),
    tar_terra_rast(
        r_big,
        c(r, r + 100, r * 10, r / 2),
        memory = "transient"
    ),
    tar_terra_tiles(
        tiles,
        raster = r_big,
        tile_fun = tile_blocksize,
        description = "split raster into tiles"
    ),
    tar_terra_rast(
        tiles_mean,
        app(tiles, \(x) mean(x, na.rm = TRUE)),
        pattern = map(tiles),
        description = "some computationaly intensive task performed on each tile"
    ),
    tar_terra_rast(
        merged_mean,
        merge(sprc(tiles_mean)),
        description = "merge tiles into a single SpatRaster"
    )
)
```

```{r}
#| label: tar_make2
#| eval: true
#| echo: true

tar_make()
```

We can see from `tar_make()` output above and the plots below that `tiles` and `tiles_mean` are both patterns with four branches each.

```{r}
#| label: tiled-plot
#| eval: true
#| echo: true
library(terra)
tar_load(tiles_mean)
op <- par(mfrow = c(2,2))
for (i in seq_along(tiles_mean)) {
    plot(tiles_mean[[i]])
}
par(op)
```

And combined, they make the full plot again.

```{r}
#| label: merged-plot
#| eval: true
#| echo: true
plot(tar_read(merged_mean))
```
