---
title: "Using `terra` with `geotargets`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using `terra` with `geotargets`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# With the root.dir option below,
# this vignette runs the R code in a temporary directory
# so new files are written to temporary storage
# and not the user's file space.
knitr::opts_knit$set(root.dir = fs::dir_create(tempfile()))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
if (identical(Sys.getenv("NOT_CRAN", unset = "false"), "false")) {
  knitr::opts_chunk$set(eval = FALSE)
}
library(targets)
library(geotargets)
```

```{r setup}
library(geotargets)
```

`geotargets` extends targets to work with geospatial dataformats, such as rasters and vectors (e.g., shapefiles).
In particular, `geotargets` aims to support use of `terra`objects in `targets` pipelines.

The design of `geotargets` is to specify target factorieslike so: `tar_<pkg>_<type>`.

In this vignette we will demonstrate the use of the `terra` R package, and we will demonstrate how to build raster (`rast`), vector (`vect`), and raster collections (`sprc`):

-   `tar_terra_rast()`
-   `tar_terra_vect()`
-   `tar_terra_sprc()`

## `tar_terra_rast()`: targets with `terra` rasters

```{r}
#| label: tar_script_rast
#| eval: true
#| echo: false

targets::tar_script({
    # contents of _targets.R:
    library(targets)
    library(geotargets)
    tar_option_set(packages = "terra")
    geotargets_option_set(gdal_raster_driver = "COG")
    list(
        tar_target(
            tif_file,
            system.file("ex/elev.tif", package = "terra"),
            format = "file"
        ),
        tar_terra_rast(
            r,
            {
                rast <- rast(tif_file)
                units(rast) <- "m"
                rast
            }
        ),
        tar_terra_rast(
            r_agg,
            aggregate(r, 2)
        )
    )
})
```


```{r}
#| label: _targets-rast
#| eval: false
#| echo: true
# contents of _targets.R:
library(targets)
library(geotargets)
tar_option_set(packages = "terra")
geotargets_option_set(gdal_raster_driver = "COG")
list(
    tar_target(
        tif_file,
        system.file("ex/elev.tif", package = "terra"),
        format = "file"
    ),
    tar_terra_rast(
        r,
        {
            rast <- rast(tif_file)
            units(rast) <- "m"
            rast
        }
    ),
    tar_terra_rast(
        r_agg,
        aggregate(r, 2)
    )
)
```

Above is a basic example showing the use of `tar_terra_rast()` in a targets pipeline. 
The command for `tar_terra_rast()` can be any function that returns a `SpatRaster` object.
In this example, we’ve set the output to a cloud optimized geotiff ("COG"), but any GDAL driver that works with `terra::writeRaster()` should also work here.
You can also set this option on a target-by-target basis with the `filetype` argument to `tar_terra_rast()`.

Running the pipeline:

```{r}
#| label: make-rast
#| eval: true
#| echo: true

tar_make()
tar_read(r)
tar_read(r_agg)
```

### Raster metadata

You may have noticed that the units for the `r` target above have gone missing.
This is due to limitations of `terra` and `targets`---`terra` saves some metadata in "sidecar" aux.json files and `targets` enforces a strict one file per target rule.
You can get around this by setting `preserve_metadata = "zip"` in `tar_terra_rast()` to save the output files, including the metadata, as a minimally compressed zip archive.
You can also set this for all raster targets with `geotargets_option_set(terra_preserve_metadata = "zip")`. 

Note: there are likely performance costs associated with this option.  
As an alternative, you can encode information in the layer names by setting `names(r) <-` which are retained even with the default `preserve_metadata = "drop"`.

```{r}
#| label: tar_script_rast_metadata
#| eval: true
#| echo: false
targets::tar_script({
    # contents of _targets.R:
    library(targets)
    library(geotargets)
    tar_option_set(packages = "terra")
    geotargets_option_set(gdal_raster_driver = "COG")
    list(
        tar_target(
            tif_file,
            system.file("ex/elev.tif", package = "terra"),
            format = "file"
        ),
        tar_terra_rast(
            r,
            {
                rast <- rast(tif_file)
                units(rast) <- "m"
                rast
            },
            preserve_metadata = "zip"
        )
    )
})

```

```{r}
#| label: _targets_rast_metadata
#| eval: false
#| echo: true
# contents of _targets.R:
library(targets)
library(geotargets)
tar_option_set(packages = "terra")
geotargets_option_set(gdal_raster_driver = "COG")
list(
    tar_target(
        tif_file,
        system.file("ex/elev.tif", package = "terra"),
        format = "file"
    ),
    tar_terra_rast(
        r,
        {
            rast <- rast(tif_file)
            units(rast) <- "m"
            rast
        },
        preserve_metadata = "zip"
    )
)
```


```{r}
#| label: make_rast_metadata
#| eval: true
#| echo: true

tar_make()
terra::units(tar_read(r))
```


## `tar_terra_vect()`: targets with terra vectors

For `terra` `SpatVector` objects, use `tar_terra_vect()` in the pipeline.  You can set vector specific options with `geotargets_option_set()` or with the `filetype` and `gdal` arguments to individual `tar_terra_vect()` calls.

```{r}
#| label: tar_script_vect
#| eval: true
#| echo: false

targets::tar_script({
    # contents of _targets.R:
    library(targets)
    library(geotargets)
    geotargets_option_set(gdal_vector_driver = "GeoJSON")
    list(
        tar_target(
            vect_file,
            system.file("ex", "lux.shp", package = "terra"),
            format = "file"
        ),
        tar_terra_vect(
            v,
            terra::vect(vect_file)
        ),
        tar_terra_vect(
            v_proj,
            terra::project(v, "EPSG:2196")
        )
    )
})
```



```{r}
#| label: _targets_vect
#| eval: false
#| echo: true
# contents of _targets.R:
library(targets)
library(geotargets)
geotargets_option_set(gdal_vector_driver = "GeoJSON")
list(
    tar_target(
        vect_file,
        system.file("ex", "lux.shp", package = "terra"),
        format = "file"
    ),
    tar_terra_vect(
        v,
        terra::vect(vect_file)
    ),
    tar_terra_vect(
        v_proj,
        terra::project(v, "EPSG:2196")
    )
)

```

```{r}
#| label: make_vect
#| eval: true
#| echo: true

tar_make()
tar_read(v)
tar_read(v_proj)
```


## `tar_terra_sprc()`: targets with terra raster collections

```{r}
#| label: tar-terra-sprc
#| eval: false
targets::tar_dir({ # tar_dir() runs code from a temporary directory.
  library(geotargets)
  targets::tar_script({
    elev_scale <- function(z = 1, projection = "EPSG:4326") {
      terra::project(
        terra::rast(system.file("ex", "elev.tif", package = "terra")) * z,
        projection
      )
    }
    list(
      tar_terra_sprc(
        raster_elevs,
        # two rasters, one unaltered, one scaled by factor of 2 and
        # reprojected to interrupted good homolosine
        command = terra::sprc(list(
          elev_scale(1),
          elev_scale(2, "+proj=igh")
        ))
      )
    )
  })
  targets::tar_make()
  x <- targets::tar_read(raster_elevs)
})
```
